\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{{#1}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Justin Le},
            pdftitle={A Non-Unique Monad Instance},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% Make links footnotes instead of hotlinks:
\renewcommand{\href}[2]{#2\footnote{\url{#1}}}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\title{A Non-Unique Monad Instance}
\author{Justin Le}
\date{January 28, 2015}

\begin{document}
\maketitle

\emph{Originally posted on
\textbf{\href{http://mstksg.github.com/inCode/}{in Code}}.}

Just stopping in for a short post before continuing with a long-overdue
series or two :) This post is a bit of a short fun one that describes a
quest I had, and hopefully some useful extra ideas I found along the
way.

Soon after I discovered Haskell, one question has plagued my mind. Day
and night, I wondered\ldots{}

\begin{quote}
Are there any Haskell types with more than one unique \texttt{Monad}
instance?
\end{quote}

This was a question that was pretty simple\ldots{}so simple that I was
sure many people had already asked and answered this. But I couldn't
really find any answers and nobody I asked at the time could really give
me one either, so this soon embedded itself as a pretty deep mystery to
my psyche.

The background?

\section{Functor and Applicative}\label{functor-and-applicative}

All \texttt{Functor} instances, if they exist, are \emph{unique} for the
type. The type uniquely determines the instance. There is only one
possible \texttt{Functor} instance for \texttt{{[}{]}}, one possible
\texttt{Functor} instance for \texttt{Maybe}, \texttt{Either}, etc.

This fact is taken advantage of by GHC to allow you to derive, for some
types, a \texttt{Functor} instance automatically.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\FunctionTok{>} \FunctionTok{:}\NormalTok{set }\FunctionTok{-}\DataTypeTok{XDeriveFunctor}
\NormalTok{ghci}\FunctionTok{>} \KeywordTok{data} \DataTypeTok{Foo} \NormalTok{a }\FunctionTok{=} \DataTypeTok{Bar} \NormalTok{[a] (}\DataTypeTok{Maybe} \NormalTok{(}\DataTypeTok{Foo} \NormalTok{a)) }\FunctionTok{|} \DataTypeTok{Baz} \NormalTok{(}\DataTypeTok{Either} \DataTypeTok{String} \NormalTok{a) (}\DataTypeTok{Foo} \NormalTok{a)}
\NormalTok{ghci}\FunctionTok{>} \KeywordTok{let} \NormalTok{x }\FunctionTok{=} \DataTypeTok{Bar} \NormalTok{[}\DecValTok{1}\NormalTok{, }\DecValTok{3}\NormalTok{] (}\DataTypeTok{Just} \NormalTok{(}\DataTypeTok{Baz} \NormalTok{(}\DataTypeTok{Right} \DecValTok{4}\NormalTok{) (}\DataTypeTok{Bar} \NormalTok{[}\DecValTok{10}\NormalTok{] }\DataTypeTok{Nothing}\NormalTok{)))}
\NormalTok{ghci}\FunctionTok{>} \NormalTok{fmap (}\FunctionTok{*}\DecValTok{2}\NormalTok{) x}
\DataTypeTok{Bar} \NormalTok{[}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{] (}\DataTypeTok{Just} \NormalTok{(}\DataTypeTok{Baz} \NormalTok{(}\DataTypeTok{Right} \DecValTok{8}\NormalTok{) (}\DataTypeTok{Bar} \NormalTok{[}\DecValTok{20}\NormalTok{] }\DataTypeTok{Nothing}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

There is no other possible \texttt{Functor} instance for that data type.
Go ahead, try :D

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Foo} \NormalTok{a }\FunctionTok{=} \DataTypeTok{Bar} \NormalTok{[a] (}\DataTypeTok{Maybe} \NormalTok{(}\DataTypeTok{Foo} \NormalTok{a)) }\FunctionTok{|} \DataTypeTok{Baz} \NormalTok{(}\DataTypeTok{Either} \DataTypeTok{String} \NormalTok{a) (}\DataTypeTok{Foo} \NormalTok{a)}

\KeywordTok{instance} \DataTypeTok{Functor} \DataTypeTok{Foo} \KeywordTok{where}
    \NormalTok{fmap f (}\DataTypeTok{Bar} \NormalTok{xs y) }\FunctionTok{=} \DataTypeTok{Bar} \NormalTok{(fmap f xs) (fmap f y)}
    \NormalTok{fmap f (}\DataTypeTok{Baz} \NormalTok{x fy) }\FunctionTok{=} \DataTypeTok{Baz} \NormalTok{(fmap f x) (fmap f fy)}
\end{Highlighting}
\end{Shaded}

However, this is not the case for \texttt{Applicative}. Everyone knows
of course about the normal (cartesian product) Applicative instance and
the zippy Applicative instance for list:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \NormalTok{[] }\KeywordTok{where}
    \NormalTok{pure x    }\FunctionTok{=} \NormalTok{[x]}
    \NormalTok{fs }\FunctionTok{<*>} \NormalTok{xs }\FunctionTok{=} \NormalTok{[ f x }\FunctionTok{|} \NormalTok{f }\OtherTok{<-} \NormalTok{fs, x }\OtherTok{<-} \NormalTok{xs ]}

\KeywordTok{instance} \DataTypeTok{Applicative} \NormalTok{[] }\KeywordTok{where}
    \NormalTok{pure      }\FunctionTok{=} \NormalTok{repeat}
    \NormalTok{fs }\FunctionTok{<*>} \NormalTok{xs }\FunctionTok{=} \NormalTok{zipWith (}\FunctionTok{$}\NormalTok{) fs xs}
\end{Highlighting}
\end{Shaded}

What is also fairly established is that every \emph{noncommutative}
\texttt{Applicative} instance also has a ``flipped'' version:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{-- a flipped IO Applicative}
\KeywordTok{data} \DataTypeTok{FlipIO} \NormalTok{a }\FunctionTok{=} \DataTypeTok{FlipIO} \NormalTok{\{}\OtherTok{ runFlipIO ::} \DataTypeTok{IO} \NormalTok{a \}}

\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{FlipIO} \KeywordTok{where}
    \NormalTok{pure x    }\FunctionTok{=} \DataTypeTok{FlipIO} \NormalTok{(pure x)}
    \NormalTok{fi }\FunctionTok{<*>} \NormalTok{xi }\FunctionTok{=} \DataTypeTok{FlipIO} \FunctionTok{$} \KeywordTok{do}
                  \NormalTok{x }\OtherTok{<-} \NormalTok{runFlipIO xi}
                  \NormalTok{f }\OtherTok{<-} \NormalTok{runFlipIO fi     }\CommentTok{-- note the backwards effects}
                  \NormalTok{return (f x)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{State} \NormalTok{s a }\FunctionTok{=} \DataTypeTok{State} \NormalTok{\{}\OtherTok{ runState ::} \NormalTok{s }\OtherTok{->} \NormalTok{(a, s) \}}

\CommentTok{-- the normal instance}
\KeywordTok{instance} \DataTypeTok{Applicative} \NormalTok{(}\DataTypeTok{State} \NormalTok{s) }\KeywordTok{where}
    \NormalTok{pure x    }\FunctionTok{=} \DataTypeTok{State} \FunctionTok{$} \NormalTok{\textbackslash{}s0 }\OtherTok{->} \NormalTok{(x, s0)}
    \NormalTok{fs }\FunctionTok{<*>} \NormalTok{xs }\FunctionTok{=} \DataTypeTok{State} \FunctionTok{$} \NormalTok{\textbackslash{}s0 }\OtherTok{->} \KeywordTok{let} \NormalTok{(f, s1) }\FunctionTok{=} \NormalTok{runState fs s0}
                                   \NormalTok{(x, s2) }\FunctionTok{=} \NormalTok{runState xs s1}
                               \KeywordTok{in}  \NormalTok{(f x, s2)}

\CommentTok{-- the flipped instance}
\KeywordTok{instance} \DataTypeTok{Applicative} \NormalTok{(}\DataTypeTok{State} \NormalTok{s) }\KeywordTok{where}
    \NormalTok{pure x    }\FunctionTok{=} \DataTypeTok{State} \FunctionTok{$} \NormalTok{\textbackslash{}s0 }\OtherTok{->} \NormalTok{(x, s0)}
    \NormalTok{fs }\FunctionTok{<*>} \NormalTok{xs }\FunctionTok{=} \DataTypeTok{State} \FunctionTok{$} \NormalTok{\textbackslash{}s0 }\OtherTok{->} \KeywordTok{let} \NormalTok{(x, s1) }\FunctionTok{=} \NormalTok{runState xs s0}
                                   \NormalTok{(f, s2) }\FunctionTok{=} \NormalTok{runState fs s1}
                               \KeywordTok{in}  \NormalTok{(f x, s2)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\FunctionTok{>} \NormalTok{liftA2 (,) getLine getLine}
\FunctionTok{>} \NormalTok{hello         }\CommentTok{-- asking for the first field}
\FunctionTok{>} \NormalTok{world         }\CommentTok{-- asking for the second field}
\NormalTok{(}\StringTok{"hello"}\NormalTok{, }\StringTok{"world"}\NormalTok{)}
\NormalTok{ghci}\FunctionTok{>} \NormalTok{runFlipIO }\FunctionTok{$} \NormalTok{liftA2 (,) (}\DataTypeTok{FlipIO} \NormalTok{getLine) (}\DataTypeTok{FlipIO} \NormalTok{getLine)}
\FunctionTok{>} \NormalTok{hello         }\CommentTok{-- asking for the second field}
\FunctionTok{>} \NormalTok{world         }\CommentTok{-- asking for the first field}
\NormalTok{(}\StringTok{"world"}\NormalTok{, }\StringTok{"hello"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Every non-commutative \texttt{Applicative} admits an alternative
instance where ``flipping'' the order of the ``effects'' is also a valid
\texttt{Applicative} instance. So, not \texttt{Maybe} or
\texttt{Either}, but \texttt{State}, \texttt{{[}{]}}, and \texttt{IO}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{-- free "flipped" Applicative instance}
\KeywordTok{data} \DataTypeTok{Flipped} \NormalTok{f a }\FunctionTok{=} \DataTypeTok{Flipped} \NormalTok{\{}\OtherTok{ runFlipped ::} \NormalTok{f a \}}

\CommentTok{-- instance where (<*>) is the same, but the order of effects is switched}
\KeywordTok{instance} \DataTypeTok{Applicative} \NormalTok{f }\OtherTok{=>} \DataTypeTok{Applicative} \NormalTok{(}\DataTypeTok{Flipped} \NormalTok{f) }\KeywordTok{where}
    \NormalTok{pure }\FunctionTok{=} \DataTypeTok{Flipped} \FunctionTok{.} \NormalTok{pure}
    \DataTypeTok{Flipped} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{Flipped} \NormalTok{x }\FunctionTok{=} \DataTypeTok{Flipped} \FunctionTok{$} \NormalTok{liftA2 (flip (}\FunctionTok{$}\NormalTok{)) x f}
\end{Highlighting}
\end{Shaded}

So cool. Types that have \texttt{Functor} instances only have one. Types
that have \texttt{Applicative} instances very often have more than one.

So, the obvious next question is\ldots{}what about \texttt{Monad}s? Is a
\texttt{Monad} instance uniquely determined by its type?

\section{Monad}\label{monad}

The answer wasn't that simple, for me. Yes, \emph{most}
\texttt{Applicative}s in the wild are non-unique, and there was a
generating rule. But not so for \texttt{Monad}s. You can't have a
\texttt{Monad} where the effects are switched, because for
\texttt{(\textgreater{}\textgreater{}=)}, you need the effects of the
first action in order to even decide what the effects of the next action
are.

I vaguely remember from my past two data types that are very similar yet
have very different \texttt{Monad} and \texttt{Applicative} instances:
(finite) lists, (infinite)
\href{http://blog.jle.im/entry/intro-to-machines-arrows-part-1-stream-and}{streams}.
From the outset, the two have almost identical structure. A
\texttt{Stream} is just a list with no \texttt{{[}{]}}/nil:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Stream} \NormalTok{a }\FunctionTok{=} \NormalTok{a }\FunctionTok{:~} \DataTypeTok{Stream} \NormalTok{a}
\end{Highlighting}
\end{Shaded}

The \texttt{Functor} instance is identical:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Functor} \DataTypeTok{Stream} \KeywordTok{where}
    \NormalTok{fmap f (x }\FunctionTok{:~} \NormalTok{xs) }\FunctionTok{=} \NormalTok{f x }\FunctionTok{:~} \NormalTok{fmap f xs}
\end{Highlighting}
\end{Shaded}

And the (only??) \texttt{Applicative} instance is the \texttt{ZipList}
instance for lists:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{Stream} \KeywordTok{where}
    \NormalTok{pure x }\FunctionTok{=} \NormalTok{x }\FunctionTok{:~} \NormalTok{pure x}
    \NormalTok{(f }\FunctionTok{:~} \NormalTok{fs) }\FunctionTok{<*>} \NormalTok{(x }\FunctionTok{:~} \NormalTok{xs) }\FunctionTok{=} \NormalTok{f x }\FunctionTok{:~} \NormalTok{(fs }\FunctionTok{<*>} \NormalTok{xs)}
\end{Highlighting}
\end{Shaded}

The \texttt{Monad} instance is however very different from that of
lists:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Monad} \DataTypeTok{Stream} \KeywordTok{where}
    \NormalTok{return x }\FunctionTok{=} \NormalTok{x }\FunctionTok{:~} \NormalTok{return x}
    \NormalTok{xs }\FunctionTok{>>=} \NormalTok{f }\FunctionTok{=} \NormalTok{join' (fmap f xs)}
      \KeywordTok{where}
\OtherTok{        join' ::} \DataTypeTok{Stream} \NormalTok{(}\DataTypeTok{Stream} \NormalTok{a) }\OtherTok{->} \DataTypeTok{Stream} \NormalTok{a}
        \NormalTok{join' ((x }\FunctionTok{:~} \NormalTok{_) }\FunctionTok{:~} \NormalTok{yss) }\FunctionTok{=} \NormalTok{x }\FunctionTok{:~} \NormalTok{join' (fmap tail' yss)}
        \NormalTok{tail' (_ }\FunctionTok{:~} \NormalTok{xs) }\FunctionTok{=} \NormalTok{xs}
\end{Highlighting}
\end{Shaded}

The \texttt{Monad} instance itself is actually interesting enough to
write about. It all revolves \texttt{join}, where \texttt{join} takes a
stream of streams and creates a stream \emph{of the diagonals}. So it
takes the first element of the first stream, the second element of the
second stream, the third element of the third stream, etc.

This is actually a special case of the \texttt{Monad} instance for all
fixed-sized ordered containers. A length 5 vector, for example, will
have the same \texttt{Applicative} and \texttt{Monad} instance as
described here: \texttt{(\textless{}*\textgreater{})} with ``zipping'',
and \texttt{join} with grabbing the diagonal of the 5-vector of
5-vectors.

This was a promising lead, but, it doesn't take \emph{too} much thought
to see that neither lists nor \texttt{Stream} are appropriate for
\emph{both} instances.

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\textbf{Aside}

In case you were wondering, here is an elaboration :D

\begin{itemize}
\item
  Fixed length vectors can't have the normal list Applicative instance
  at all, unless they are of size 0 or 1. That's because the result
  after \texttt{(\textless{}*\textgreater{})}, the resulting list's
  length is the product of the original lists. So you can forget the
  \texttt{Monad} instance, too.
\item
  Streams give you no luck, either. The easiest way to see is by
  considering the analogous \texttt{Monad} instance, where \texttt{join}
  is the straight-up concatenation.
  \texttt{m\ \textgreater{}\textgreater{}=\ return\ ==\ m} is clearly
  violated. If \texttt{m} is an infinite list, \texttt{fmap\ return}
  gives you an infinite list of infinite lists,
  ``joining''/concatenating them back will just give you an infinite
  list of the first item in \texttt{m}.

  To put succincently, for \texttt{Stream}, \texttt{concat\ ==\ head}.
\item
  Lists can have the \texttt{Applicative} instance fine, but not the
  \texttt{Monad} instance. Here we assume that zipping and ``getting the
  diagonal'' go only as ``far as possible'', and stop when one of the
  lists is too short.

  This one is a little trickier, but the weakness is when you have lists
  of lists of lists of different lengths.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\FunctionTok{>} \KeywordTok{let} \NormalTok{counterexample }\FunctionTok{=} \NormalTok{[[[}\DecValTok{1}\NormalTok{]], [[], [}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{]]]}
\NormalTok{ghci}\FunctionTok{>} \NormalTok{join counterexample}
\NormalTok{[[}\DecValTok{1}\NormalTok{], [}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{]]}
\NormalTok{ghci}\FunctionTok{>} \NormalTok{join}\FunctionTok{.} \NormalTok{join }\FunctionTok{$} \NormalTok{counterexample}
\NormalTok{[}\DecValTok{1}\NormalTok{,}\DecValTok{3}\NormalTok{]}
\NormalTok{ghci}\FunctionTok{>} \NormalTok{fmap join counterexample}
\NormalTok{[[}\DecValTok{1}\NormalTok{], []]}
\NormalTok{ghci}\FunctionTok{>} \NormalTok{join }\FunctionTok{.} \NormalTok{fmap join }\FunctionTok{$} \NormalTok{counterexample}
\NormalTok{[}\DecValTok{1}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

  For a monad, joining the inner layer and then joining it all should be
  the same as joining it all and joining it all. The order of the
  joining shouldn't count. We can see this in the more haskelly monad
  laws by noting:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ghci}\FunctionTok{>} \NormalTok{id }\FunctionTok{<=<} \NormalTok{(id }\FunctionTok{<=<} \NormalTok{id) }\FunctionTok{$} \NormalTok{counterexample}
\NormalTok{[}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{]}
\NormalTok{ghci}\FunctionTok{>} \NormalTok{(id }\FunctionTok{<=<} \NormalTok{id) }\FunctionTok{<=<} \NormalTok{id }\FunctionTok{$} \NormalTok{counterexample}
\NormalTok{[}\DecValTok{1}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

  So, dead end here.
\end{itemize}

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

So I didn't really have any leads at that point; I tried a couple of
other paths but nothing really panned out. So I shelved it for a while.

\section{Revelation}\label{revelation}

Several centuries later\footnote{More accurately, ``about a year''}, the
final revelation came as many revelations do in Haskell --- from a hint
by Edward Kmett. He pointed out something interesting regarding a
\texttt{Monad} instance that I had yet to notice:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Monoid} \NormalTok{w }\OtherTok{=>} \DataTypeTok{Monad} \NormalTok{((,) w) }\KeywordTok{where}
\end{Highlighting}
\end{Shaded}

This is the classic ``Writer'' monad instance, which is literally about
as old as monads in functional programming is.

The key is that the \texttt{Monad} instance of \texttt{(w,)} depends on
the \texttt{Monoid} instance of \texttt{w}. This is the ``log'', so to
speak. You need a \texttt{Monoid} instance in order to make the
\texttt{Monad} instance\ldots{}and the behavior of the \texttt{Monad}
instance is directly determined by the behavior of the \texttt{Monoid}
instance of \texttt{w}.

And\ldots{}\texttt{Monoid} instances in Haskell are rarely ever unique!
A different \texttt{Monoid} instance would create a very different
\texttt{Monad} instance for the same type!

So, by factoring out the dependency on an external \texttt{Monoid}
instance, you get\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{data} \DataTypeTok{Two} \NormalTok{a }\FunctionTok{=} \DataTypeTok{One} \NormalTok{a }\FunctionTok{|} \DataTypeTok{Two} \NormalTok{a}

\KeywordTok{instance} \DataTypeTok{Functor} \DataTypeTok{Two} \KeywordTok{where}
    \NormalTok{fmap f (}\DataTypeTok{One} \NormalTok{a) }\FunctionTok{=} \DataTypeTok{One} \NormalTok{(f a)}
    \NormalTok{fmap f (}\DataTypeTok{Two} \NormalTok{a) }\FunctionTok{=} \DataTypeTok{Two} \NormalTok{(f a)}
\end{Highlighting}
\end{Shaded}

and\ldots{}voila! There it is!

This type is basically equivalent to \texttt{(Bool,\ a)}. And
\texttt{Bool} has multiple \texttt{Monoid}s on it. Instead of requiring
an outside \texttt{Monoid} instance, we can encode the instance directly
into the behavior of \texttt{(\textgreater{}\textgreater{}=)}. And here
we go!

Our instances are basically the \texttt{Writer} instance for
\texttt{(Bool,\ a)}, with different \texttt{Monoid} instances for
\texttt{Bool}.

The first instance:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{Two} \KeywordTok{where}
    \NormalTok{pure }\FunctionTok{=} \DataTypeTok{One}
    \DataTypeTok{One} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{One} \NormalTok{x }\FunctionTok{=} \DataTypeTok{One} \NormalTok{(f x)}
    \DataTypeTok{One} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{Two} \NormalTok{x }\FunctionTok{=} \DataTypeTok{Two} \NormalTok{(f x)}
    \DataTypeTok{Two} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{One} \NormalTok{x }\FunctionTok{=} \DataTypeTok{Two} \NormalTok{(f x)}
    \DataTypeTok{Two} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{Two} \NormalTok{x }\FunctionTok{=} \DataTypeTok{Two} \NormalTok{(f x)}

\KeywordTok{instance} \DataTypeTok{Monad} \DataTypeTok{Two} \KeywordTok{where}
    \NormalTok{return }\FunctionTok{=} \DataTypeTok{One}
    \DataTypeTok{One} \NormalTok{x }\FunctionTok{>>=} \NormalTok{f }\FunctionTok{=} \NormalTok{f x}
    \DataTypeTok{Two} \NormalTok{x }\FunctionTok{>>=} \NormalTok{f }\FunctionTok{=} \KeywordTok{case} \NormalTok{f x }\KeywordTok{of}
                    \DataTypeTok{One} \NormalTok{y }\OtherTok{->} \DataTypeTok{Two} \NormalTok{y}
                    \DataTypeTok{Two} \NormalTok{y }\OtherTok{->} \DataTypeTok{Two} \NormalTok{y}
\end{Highlighting}
\end{Shaded}

Which represents the monoids formed by \texttt{(\&\&)} with
\texttt{True} or by \texttt{(\textbar{}\textbar{})} with \texttt{False}
(depending on which one you pick as \texttt{True} and which one you pick
as \texttt{False}; the two instances are isomorphic)

The second:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{instance} \DataTypeTok{Applicative} \DataTypeTok{Two} \KeywordTok{where}
    \NormalTok{pure }\FunctionTok{=} \DataTypeTok{One}
    \DataTypeTok{One} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{One} \NormalTok{x }\FunctionTok{=} \DataTypeTok{One} \NormalTok{(f x)}
    \DataTypeTok{One} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{Two} \NormalTok{x }\FunctionTok{=} \DataTypeTok{Two} \NormalTok{(f x)}
    \DataTypeTok{Two} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{One} \NormalTok{x }\FunctionTok{=} \DataTypeTok{Two} \NormalTok{(f x)}
    \DataTypeTok{Two} \NormalTok{f }\FunctionTok{<*>} \DataTypeTok{Two} \NormalTok{x }\FunctionTok{=} \DataTypeTok{One} \NormalTok{(f x)}

\KeywordTok{instance} \DataTypeTok{Monad} \DataTypeTok{Two} \KeywordTok{where}
    \NormalTok{return }\FunctionTok{=} \DataTypeTok{One}
    \DataTypeTok{One} \NormalTok{x }\FunctionTok{>>=} \NormalTok{f }\FunctionTok{=} \NormalTok{f x}
    \DataTypeTok{Two} \NormalTok{x }\FunctionTok{>>=} \NormalTok{f }\FunctionTok{=} \KeywordTok{case} \NormalTok{f x }\KeywordTok{of}
                    \DataTypeTok{One} \NormalTok{y }\OtherTok{->} \DataTypeTok{Two} \NormalTok{y}
                    \DataTypeTok{Two} \NormalTok{y }\OtherTok{->} \DataTypeTok{One} \NormalTok{y}
\end{Highlighting}
\end{Shaded}

Which represents the monoid formed by \texttt{(/=)} (or ``XOR'') with
\texttt{False}.

And there you go. One type, two possible unique, non-isomorphic
\texttt{Monad} instances.

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\textbf{Aside}

One interesting thing to note is that the Monad instance for
\texttt{(-\textgreater{})\ a} requires no monoid constraint, and the
Monad instance for \texttt{(,)\ a} \emph{does}.

Interestingly enough, if we look at \emph{comonads}, the Comonad
instance for \texttt{(-\textgreater{})\ a} \emph{does} require a monoid
constraint on \texttt{a} (so for example there are many unique Comonad
instances for things isomorphic to \texttt{(-\textgreater{})\ a} where
\texttt{a} has more than one Monoid instance) and and the Comonad
instance for \texttt{(,)\ a} \emph{does not} require a monoid constraint
on \texttt{a}.

Is there some duality at play here?

The answer is, apparently, yes! But according to Edward Kmett, it is one
that is pretty hard to arrive at and a big headache and overall not
worth the time to dig into. So you're going to have to take my
second-hand word for it.

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\end{document}
